### 2013-03-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updates to spn equations introduction

The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, very large, or both, requiring tremendous computational
resources to generate an adequate solution. Modern deterministic
methods for large scale problems are commonly variants on the discrete
ordinates ($S_N$) method \citep{evans_denovo:_2010}. For full reactor
core neutronics simulations, the $S_N$ method requires potentially
trillions of unknown angular flux moments to be computed to achieve
good accuracy for the repsonses of interest. Other forms of the
transport problem, including the $P_N$ method, take on a simpler form
than the more common $S_N$ methods but lack in accuracy when compared
while still requiring considerable computational resources for
solutions in multiple dimensions.

In the 1960's, Gelbard developed an ad-hoc multidimensional extension
of the simple single dimension planar $P_N$ equations that created a
system of coupled, diffusion-like equations known as the simplified
$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
methods was either widely unknown, widely unused, or combination of
both even though numerical studies showed promising results with
better solutions than diffusion theory and a significant reduction in
computational time over more accurate methods such as discrete
ordinates. Why did this happen? A significant problem, pointed out by
Larsen, was that little rigor had been applied to the formulation of
the $SP_N$ equations since their derivation through primarily
heuristic arguments. Instead, studies at that time focused on simply
comparing the results of the method to other contemporary transport
solution strategies. In addition, many problems of interest from the
literature at the time were either solved using nodal-type methods for
reactor-sized problems or $S_N$-type methods for benchmark problems
with intricate material configurations and potentially large flux
gradients over small spatial domains.

@@

To further motivate moving in this direction, recent developments in
the Exnihilo neutronics package at Oak Ridge National Laboratory have
permitted generation of the $SP_N$ system of equations for detailed
full core reactor models. By fully forming these equations and
formulating them as a linear algebra problem, we now have access to
all of the modern advancements in computational linear algebra
including Krylov solvers for asymmetric systems and preconditioning
methods such as algebraic multigrid. This leads us to then explore the
applicability of our work in discrete Monte Carlo methods for linear
systems as a possible solution method for the $SP_N$ equations. If
formulated correctly, we hypothesize that signficant improvement in
usage of computational resources may be observed compared to modern
solution techniques such as those suggested. In addition, solving the
$SP_N$ equations in this way also breaks away from the $S_N$ forms of
parallelism where spatial parallelism is achieved by an efficient
parallel sweep, angular efficiency achieved by pipelining, and energy
parallelism achieved by decoupling the groups. With the $SP_N$
equations as a full matrix system, we now can parallelize the problem
as prescribed by the linear solver, which may be significantly more
scalable than current $S_N$ transport practices.

In this chapter, we derive the $SP_N$ equations, closely following the
work of Evans. We begin by stating the general time-independent
neutron transport equation followed by a derivation of the $P_N$
equations in planar geometry for multiple energy groups. From these
equations, we then apply a set of approximations to yield the
multi-dimensional, multi-group $SP_N$ equations for fixed source
problems.

>>>>

The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, very large, or both, requiring tremendous computational
resources to generate an adequate solution. Modern deterministic
methods for large scale problems are commonly variants on the discrete
ordinates ($S_N$) method \citep{evans_denovo:_2010}. For full reactor
core neutronics simulations, the $S_N$ method requires potentially
trillions of unknown angular flux moments to be computed to achieve
good accuracy for the repsonses of interest
\citep{slaybaugh_acceleration_2011}. Other forms of the transport
problem, including the $P_N$ method, take on a simpler form than the
more common $S_N$ methods but lack in accuracy when compared while
still requiring considerable computational resources for solutions in
multiple dimensions.

In the 1960's, Gelbard developed an ad-hoc multidimensional extension
of the simple single dimension planar $P_N$ equations that created a
system of coupled, diffusion-like equations known as the simplified
$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
method was either widely unknown, widely unused, or combination of
both even though numerical studies showed promising results with
better solutions than diffusion theory and a significant reduction in
computational time over more accurate methods such as discrete
ordinates. Why did this happen? A significant problem, pointed out by
Larsen, was that little rigor had been applied to the formulation of
the $SP_N$ equations since their derivation through primarily
heuristic arguments. Instead, studies at that time focused on simply
comparing the results of the method to other contemporary transport
solution strategies. In addition, many problems of interest from the
literature at the time were either solved using nodal-type methods for
reactor-sized problems or $S_N$-type methods for benchmark problems
with intricate material configurations and potentially large flux
gradients over small spatial domains.

@@

To further motivate moving in this direction, recent developments in
the Exnihilo neutronics package at Oak Ridge National Laboratory have
permitted generation of the $SP_N$ system of equations for detailed
full core reactor models. By fully forming these equations and
formulating them as a linear algebra problem instead of using the
explicit iterative methods of the past, we now have access to all of
the modern advancements in computational linear algebra including
Krylov solvers for asymmetric systems and preconditioning methods such
as algebraic multigrid. This leads us to then explore the
applicability of our work in discrete Monte Carlo methods for linear
systems as a possible solution method for the $SP_N$ equations. If
formulated correctly, we hypothesize that signficant improvement in
usage of computational resources may be observed compared to modern
solution techniques such as those suggested due to the form of the
matrices generated by the discretization. In addition, solving the
$SP_N$ equations in this way also breaks away from the $S_N$ forms of
parallelism where spatial parallelism is achieved by an efficient
parallel sweep, angular efficiency achieved by pipelining, and energy
parallelism achieved by decoupling the groups. With the $SP_N$
equations as a full matrix system, we now can parallelize the problem
as prescribed by the linear solver, which may be significantly more
scalable than current $S_N$ transport practices.

In this chapter, we derive the $SP_N$ equations, closely following the
work of Evans, in order to gain full understanding of the underlying
system and its behavior in a Monte Carlo context. We begin by stating
the general time-independent neutron transport equation followed by a
derivation of the $P_N$ equations in planar geometry for multiple
energy groups. From these equations, we then apply a set of
approximations to yield the multi-dimensional, multi-group $SP_N$
equations for fixed source problems. Boundary conditions are also
discussed.
